---
title: "Fixing Cluttered PCA plots"
author: "B. Valderrama"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

### Required packages 
```{r, warning=FALSE, message=FALSE}
library(tidyverse) # work with data and plotting
library(patchwork) # 'paste' plot together
library(ggnewscale) # to use more than one color scale in a plot
```


### Generating data

```{r}
set.seed(1997)

group1 <- data.frame(PC1 = rnorm(1000, mean = 0.5),
                     PC2 = rnorm(1000),
                     group = "group_1")

group2 <- data.frame(PC1 = rnorm(1000, mean = -0.5),
                     PC2 = rnorm(1000, mean = 0.5, sd = 0.75),
                     group = "group_2")

group3 <- data.frame(PC1 = rnorm(1000, sd = 1.5),
                     PC2 = rnorm(1000, mean = -0.5),
                     group = "group_3")

data <- rbind(group1, group2, group3)

# mix the observations to avoid that groups cover each other
data <- data[sample(1:nrow(data), size = nrow(data)), ]

head(data)
```

# Plotting

### Cluttered PCA

```{r, fig.align='center'}
basic <- data %>% 
        ggplot(aes(x = PC1, y = PC2, fill = group)) +
        geom_point(shape = 21, size = 2, alpha = 0.75) +
        scale_fill_viridis_d() + # custom colors
        theme_minimal() +
        theme(legend.text = element_text(size = 10),
              legend.title = element_text(size = 12))

basic 
```

### Box or density plots

One alternative commonly used in the microbiome literature is to add plots on
the top and side of the PCA to show the distribution of the groups across the 
axes. Here I show box plots, but density plots are another common choice. 

```{r, fig.align='center', warning=FALSE, message=FALSE}
library(ggside)

pca_with_bp <- basic +
        geom_xsideboxplot(orientation = "y", 
                          outliers = FALSE, 
                          show.legend = FALSE) +
        geom_ysideboxplot(orientation = "x",
                          outliers = FALSE, 
                          show.legend = FALSE) +
        theme_ggside_void(); pca_with_bp
```

### Binning PCA space

```{r, fig.align='center'}
binning_pca <- function(group){
        
        # data of the group of interest 
        group_df <- data %>% 
                filter(group == {{group}})
        
        data %>% 
                ggplot(aes(x = PC1, y = PC2)) +
                # we add the background with all the observations in grey
                geom_bin_2d(bins = 20) +
                scale_fill_gradient(low = "grey50", high = "grey50") +
                
                # we add the data of the group of interest on top
                new_scale("fill") +
                geom_bin_2d(data = group_df, bins = 20) +
                scale_fill_continuous(type = "viridis") +
                
                # add a title to identify the group
                labs(title = gsub(x = group, pattern = "_", replacement = " ")) +
                
                coord_fixed() +
                
                theme_minimal() +
                theme(legend.position = "none")
}

binned_pcas <- lapply(c("group_1", "group_2", "group_3"), binning_pca) %>% 
        wrap_plots(nrow = 1); binned_pcas
```

## Show alternatives to the cluttered PCA

```{r, fig.align='center'}

first_row <- ((basic | pca_with_bp) & theme(legend.position = "none"))
 
first_row / binned_pcas

# export figures
ggsave(filename = "../figures/alternatives_to_cluttered_pca.png", height = 4.5, width = 6)
ggsave(filename = "../figures/alternatives_to_cluttered_pca.svg", height = 4.5, width = 6)

```